How residence time constrains diversity

### 2015-06-20 paper/Manuscript.Rmd
finishing up Results -KL

**P1**—Biological and ecological change is marked by time. From rates of metabolism, growth, and evolution to the duration of seasons, succession, and life history, time is a primary dimension of biology. But despite its universal relevance, time is more often used as an independent variable or to characterize rates of change, than it is treated as a primary driver of change and biodiversity *per se*. The general influence of time  is so rarely mentioned that it may be self-evident, non-existent, or taken for granted. This is unlike the influence of other general variables, such as space and abundance.

**P2**—Law-like patterns such as the species-area relationship and the species-abundance distribution are central to the ecological study of biodiversity (Hubbell 2001, McGill 2010, Harte 2011). Yet, a similarly general time-related pattern (species-time relationship) is rarely studied and is neither central to nor predicted by general ecological theory (White, Adler et al.). While area and abundance are common inputs to biodiversity theories (McGill 2010), none use time as an input. And, while at least one popular ecological subfield is devoted to spatial dynamics, i.e., spatial ecology (16000 results on Google scholar), "temporal ecology" has yet to be considered a subfield (590 results on Google scholar) and if temporal dynamics are often along with space, i.e., "spatio-temporal".

**P3**—The most general spatial principle is very simple, "near things are more similar than far things" (Tobler's Law) and even a precise knowledge of how abundance constrains diversity (cites), there is no general ecological rule or law relating to time. Unlike area, distance, abundance, and even the number of species (i.e., richness), there are no studies on the general influence of time on biodiversity. Other than the species-time relationship, there are no general time-related ecological patterns. Yet, at least one aspect of time applies to all components of any biological or ecological system, i.e., the amount of time that an individual organism, an amount of resource, a species, etc., remains in a system, i.e., residence time.

**P4**—Put simply, viable populations are maintained when individuals can reproduce before dying or emigrating but not so quickly that resource consumption exhausts the rate of supply. In this way, the timing of life history and population growth is constrained by residence time or even optimized for it (cite). This principle underpins the theory of the chemostat and makes residence time a master variable for controlling growth and biomass in experimental microcosms and engineered bioreactors. It also makes residence time among the most important variables influencing stability, function, and biodiversity in estuaries (cites). Yet, the influence of residence time is poorly known outside aquatic systems and though it is often manipulated to constrain growth and biomass, residence time has yet been framed as a general time-related constraint in any system, whether aquatic, terrestrial, or host-related. Yet, while many organisms are at the mercy of currents, all organisms are at the mercy of time.

@@

**Residence time predictions**

@@

@@

>>>>

**P1**—Biological and ecological change is marked by time and from rates of metabolism, growth, and evolution to the duration of seasons, succession, and life history, time is the fundamental dimension of biology. But despite its universal relevance in biological and ecological systems, time is more often used as an independent variable or to characterize rates of change, than it is treated as a primary driver of change and biodiversity. While the large lack of study into the general influence of time is recognized among some ecologists (cite), the precise importance or general influence of time is so rarely mentioned that it may be self-evident, non-existent, or commonly taken for granted. But, this is unlike the influence of other general variables such as space and abundance.

**P2**—Law-like patterns of space and abundance such as the species-area relationship and the species-abundance distribution are central to the ecological study of biodiversity (Hubbell 2001, McGill 2010, Harte 2011). Yet, a similarly general temporal pattern (species-time relationship) is rarely studied and is neither central to nor predicted by general ecological theory (White, Adler et al.). While area and abundance are common inputs to biodiversity theories (McGill 2010), no general theory uses time *per se* as an input. And, while at least one popular ecological subfield is devoted to spatial dynamics, i.e., spatial ecology (16000 results on Google scholar), "temporal ecology" has yet to be deemed a subfield (590 results on Google scholar) while temporal dynamics are studied along with spatial, i.e., "spatio-temporal".

**P3**—Even if the general influence of time were very simple, it would be no simpler and no less profound than the most general spatial principle, i.e., "near things are more similar than far things", also known as Tobler's Law or Tobler's First Rule of Geography (cite). Likewise, it might be no less straightforward, understated, and mathematically necessary than the constraining influence of total abundance and sample size on intensively studied patterns of commonness and rarity (Locey and White 2013, Xiao et al. 2015). Yet there is no general ecological rule or law relating to time. There are no studies on the general influence of time on biodiversity and, other than the species-time relationship, there are no general time-related ecological patterns. Yet, at least one aspect of time applies to all components of any biological or ecological system, i.e., the amount of time that an individual organism, resource particle, species, etc., remains in a system; residence time.

**P4**—Put simply, viable populations are maintained when individuals can reproduce before dying or emigrating but not so quickly that resource consumption exhausts the rate of supply. In this way, the timing of life history and population growth is constrained by residence time or perhaps even optimized for it (cite). This principle underpins chemostat theory, making residence time a master variable for controlling growth and biomass in experimental microcosms and engineered bioreactors. It also makes residence time among the most important variables influencing stability, function, and biodiversity in sensitive ecosystems like estuaries (cites). Yet, the influence of residence time is poorly known outside aquatic systems and though it is often manipulated to control growth and biomass, residence time has yet been framed as a time-related constraint common to any system, whether aquatic, terrestrial, or host-related. Yet, while many organisms are at the mercy of currents, all organisms are at the mercy of time.

@@

**Primary residence time predictions that should hold up to ecological complexity**

@@

**Residence time constrains *N*, while *N* constrains commonness and rarity**

@@

**Residence time influences at least one general life history tradeoff**

*Growth rate vs. maintenance costs—*Organisms with slow growth should not be favored in systems of fast turnover, while organisms with high maintenance costs should not be favored in systems of slow resource supply. Yet, unless growth rate and maintenance are directly related (high growth requiring high maintenance), the relationship between these two traits could be tenuous.
